---
title: "R workbook for Geospatial Health Data by Paula Moraga"
author: "Mathupanee Oonsivilai"
date: "2024-05-16"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
This notebook is not meant to be knitted. It was meant as a code tutorial for the DriaDD group bookclub for Geospatial Health Data by Paula Moraga (2019) which can be found at https://www.paulamoraga.com/book-geospatial/. 

This notebook adds: a reference to the R Docker container which will run code out-of-the-box, renv version lock on packages, installation guidepost for missing packages, some contextual notes and links.

## Environment setup via renv

rgdal last frozen at 2023-06-01.
Latest relevant docker image before that freeze date is rocker/geospatial:4.3.1.
Thus this notebook is developed on rocker/geospatial:4.3.1 (731d51d434739f72f930acad26456a20ef42cd1209dc53a21aae7138bd36ee2a)

Just install the packages, then check renv version lock only if that fails.

Check status of renv. 
```{r check status}
# It acts a bit wonky with rocker, but it's just good to have package version listing.
#renv::init()
#file.remove(".Rprofile")
#unlink("renv", recursive=TRUE)
```

Need to install these packages on container creation.
```{r install missing packages, message=FALSE}
install.packages("cholera")
devtools::install_version("rgdal", version = "1.6-7", repos = "http://cran.us.r-project.org")
install.packages("leaflet")
install.packages("webshot")
install.packages("mapview")
install.packages("leaflet.extras2")
```

```{r load useful libraries, message=FALSE}
library(tidyverse)
```

## Chapter 2: Spatial data and R packages for mapping

### Example Geodata Types
```{r Areal Data-SIDS North Carolina, echo=FALSE}
library(sf)
library(ggplot2)
library(viridis)
nc <- st_read(system.file("shape/nc.shp", package = "sf"), quiet=TRUE)
test_nc <- nc %>% dplyr::select(SID74)#, geometry) # Only need these 2 columns to function
ggplot(data = test_nc, aes(fill = SID74)) + geom_sf() +
  scale_fill_viridis() + theme_bw()
```

```{r Geostat Data-Rainfall Parana Brazil, echo=FALSE}
library(geoR)
pb <- data.frame(cbind(parana$coords, Rainfall = parana$data))
pb_border <- data.frame(parana$border)
ggplot(pb) + geom_point(aes(east, north, color=Rainfall), size=2) +
  coord_fixed(ratio = 1) +
  scale_color_gradient(low = "blue", high="orange") +
  geom_path(data = pb_border, aes(east, north)) +
  theme_bw()
```

```{r Point Pattern Data-Cholera}
library(cholera)
rng <- mapRange() #specific function to return xlim and ylim of Snow's map
snow <- fatalities
plot(fatalities[, c("x", "y")],
     pch = 15, col = "black",
     cex = 0.5, xlim = rng$x, ylim = rng$y, asp=1,
     frame.plot = FALSE, axes = FALSE, xlab = "", ylab = ""
     )
addRoads() #specific function to add roads to plot
```

```{r Show structure of different types of data}
str(nc)
str(pb)
str(snow)
```

### Coordinate Reference Systems (CRS)

prog4 strings maybe outdated. https://inbo.github.io/tutorials/tutorials/spatial_crs_coding/
```{r example PROJ4 strings}
# WGS84
"+proj=longlat +ellps=WGS84 +datum=WGS84 +no_defs"
# UTM zone 29
"+proj=utm +zone=29 +ellps=WGS84 +datum=WGS84 +units=m +no_defs"
# UTM zone 29 south
"+proj=utm +zone=29 +ellps=WGS84 +datum=WGS84 +units=m +no_defs +south"
```

rdgal package has been retired https://r-spatial.org/r/2022/04/12/evolution.html

```{r show all available CRS in R}
View(rgdal::make_EPSG())
```

```{r example conversion between projections}
library(rgdal)

# create data with coordinates given by longitude and latitude
d <- data.frame(long = rnorm(100, 0, 1), lat = rnorm(100, 0, 1))
coordinates(d) <- c("long", "lat")

# assign CRS WGS84 longitude/latitude
proj4string(d) <- CRS("+proj=longlat +ellps=WGS84
                      +datum=WGS84 +no_defs")

# reproject data from longitude/latitude to UTM zone 35 south
d_new <- spTransform(d, CRS("+proj=utm +zone=35 +ellps=WGS84
                      +datum=WGS84 +units=m +no_defs +south"))

# add columns UTMx and UTMy
d_new$UTMx <- coordinates(d_new)[, 1]
d_new$UTMy <- coordinates(d_new)[, 2]
```

### Shapefiles

```{r read in files with readOGR example}
# name of the shapefile of North Carolina of the sf package
nameshp <- system.file("shape/nc.shp", package = "sf")
# read shapefile with readOGR()
library(rgdal)
map <- readOGR(nameshp, verbose = FALSE)

class(map)
```

```{r readOGR cont. 1}
head(map@data)
```

```{r readOGR cont. 2}
plot(map)
```

```{r Read in with st_read() example}
# read shapefile with st_read()
library(sf)
map <- st_read(nameshp, quiet = TRUE)

class(map)
```


```{r st_read cont. 1}
head(map)
```

```{r st_read cont. 2}
plot(map)
```

### Making maps with R

```{r ggplot2 via geom_sf example}
library(sf)
library(ggplot2)
nameshp <- system.file("shape/nc.shp", package = "sf")
map <- st_read(nameshp, quiet = TRUE)
ggplot(map) + geom_sf(aes(fill = SID74)) + theme_bw()
```

```{r ggplot2 change color example}
library(viridis)
map <- st_as_sf(map)
ggplot(map) + geom_sf(aes(fill = SID74)) +
  scale_fill_viridis() + theme_bw()
```
```{r ggplot2 save plot example}
png("plot.png")
ggplot(map) + geom_sf(aes(fill = SID74)) +
  scale_fill_viridis() + theme_bw()
dev.off()
```

```{r transform EPSG4267 to EPSG4326 as required by leaflet}
st_crs(map) # Originally NAD27, need WGS84
map <- st_transform(map, 4326)
```

```{r leaflet example}
library(leaflet)

pal <- colorNumeric("YlOrRd", domain = map$SID74)

leaflet(map) %>%
  addTiles() %>%
  addPolygons(
    color = "white", fillColor = ~ pal(SID74),
    fillOpacity = 0.5
  ) %>%
  addLegend(pal = pal, values = ~SID74, opacity = 1)
```

```{r leaflet saveWidget example}
library(htmlwidgets)
p<- leaflet(map) %>%
  addTiles() %>%
  addPolygons(
    color = "white", fillColor = ~ pal(SID74),
    fillOpacity = 1
  ) %>%
  addLegend(pal = pal, values = ~SID74, opacity = 1)
saveWidget(p, "examplewidget.html")
```

```{r webshot example (not working)}
# Some sort of issue with OpenSSL version and whatnot. https://stackoverflow.com/questions/73004195/phantomjs-wont-install-autoconfiguration-error
library(webshot)
webshot("http://www.google.com")
```

```{r mapview example}
library(mapview)
mapview(map, zcol = "SID74")
```

```{r mapview customize background map example}
library(RColorBrewer)
pal <- colorRampPalette(brewer.pal(9, "YlOrRd"))
mapview(map,
  zcol = "SID74",
  map.types = "CartoDB.DarkMatter",
  col.regions = pal
)
```


```{r leafsync sync multiple maps}
library(leafsync)
m74 <- mapview(map, zcol = "SID74")
m79 <- mapview(map, zcol = "SID79")
m <- leafsync::sync(m74, m79)
m
```

Original code did not sync legends, but found example code that did in Moraga's other book Spatial Statistics for Data Science (2023).
```{r leafsync sync multiple maps improved}
library(leafsync)
pal <- colorRampPalette(brewer.pal(9, "YlOrRd"))
at <- seq(min(c(map$SID74, map$SID79)), max(c(map$SID74, map$SID79)),
          length.out = 8)
map_type <- "CartoDB.Positron"
m74 <- mapview(map, zcol = "SID74", map.types = map_type, col.regions = pal, at = at)
m79 <- mapview(map, zcol = "SID79", map.types = map_type, col.regions = pal, at = at)
m <- leafsync::sync(m74, m79)
m
```
```{r leafsync sync multiple maps slider via leaflet.extras2}
library(leaflet.extras2)
m74 | m79
```



```{r tmap example}
library(tmap)
tmap_mode("view") #interactive
#tmap_mode("plot") #static
tm_shape(map) + tm_polygons("SID74")
```

### Update from PROJ4 string to WKT2 strings  with sf

Running examples from link in note about outdated PROJ4 strings.
```{r load in example data from rgdal}
library(sp)
library(raster)
library(sf)

cities <- 
  st_read(system.file("vectors/cities.shp", package = "rgdal"))

```

```{r strip to bare XY coords and manually get WKT2 for WGS82}
# Convert to plain data frame with XY coordinates to add CRS ourselves
cities <- 
  cbind(st_drop_geometry(cities), 
        st_coordinates(cities))
head(cities, 10) # top 10 rows

# Get WKT2 string representing WGS84
crs_wgs84 <- st_crs(4326) # WGS 84 has EPSG code 4326
```

```{r convert data to sf object}
# Convert to sf object
cities2 <- st_as_sf(cities, coords = c("X", "Y"))
st_crs(cities2)
```

```{r add CRS to data}
# Add CRS 
st_crs(cities2) <- 4326
st_crs(cities2)
```

sp package has been outgrown by sf package.
raster package is superseded by terra package.
Link goes through examples of how to do conversion in sp and raster.
